Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-12 icfp.tex
RM Edits Sec 2. prelude

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is
well-typed under a linear type system intrinsically depends on the
definition of \emph{consuming} a resource.
In general, \emph{consuming} a resource is equated with its 
\emph{syntactic occurrence} in the program. However, as this section
makes clear, when considering linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
Consider the following program in a functional Haskell-like language,
where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
\lstinline{handle} is returned:
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%

In the code above, the \lstinline{handle} appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e.,~evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with its syntactic occurrences, since depending on the evaluation strategy
of the language the computation that closes the handle might not be
executed, and if so, the \lstinline{handle} in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under
under call-by-need semantics.  %With \emph{eager}
% evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
The let
bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by
calling \lstinline{f}.
Under the lens of \emph{call-by-need}, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated with \emph{call-by-value}, where
the let-bound expression is always evaluated.

Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively computed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
%
From this observation, and exploring the connection between computation and evaluation,
% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
% above example and for programs in general, should be defined in function of the
% language's evaluation strategy.
%
we turn our focus to \emph{linearity} under \emph{call-by-need}, not only
because GHC Core is \emph{call-by-need}, but also because the distinction
between semantically and syntactically consuming a resource is only exposed
under \emph{non-strict} semantics.
%
% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.

@@

Aligned with our original motivation of typechecking linearity in GHC Core such
that optimising transformations preserve linearity, and with the goal
of understanding linearity in a non-strict context, this section aims to
build an intuition for semantic linearity through examples of Core
programs that are semantically linear but rejected by (the current) Core linear type
system.
% BETTER?:
%Aligned with our motivation of typechecking linearity in GHC Core such
%that optimising transformations preserve linearity, and in order to explain
%linearity in a non-strict context, we illustrate semantic linearity
%through examples of Core programs that are semantically linear but rejected by
%Core's linear type system -- building intuition towards semantic linearity in a
%lazy language.
%
In the examples, a \colorbox{working}{\workingcolorname} background highlights programs that are
syntactically linear and accepted by Core's naive linear type system. A
\colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
\emph{semantically} linear, but are not seen as linear by Core's (naive wrt laziness) linear type
system. Notably, the linear type system we develop in this work accepts all
\colorbox{notyet}{\notyetcolorname} programs.
% (while the few \colorbox{limitation}{\limitationcolorname} programs are not accepted).
A \colorbox{noway}{\nowaycolorname} background indicates that the program
violates linearity semantically, i.e.~the program effectively
discards or duplicates linear resources. We discuss the three main
syntactic constructs of Core, let bindings (non-recursive and
recursive) and case expressions.

>>>>

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is
well-typed under a linear type system intrinsically depends on the
definition of \emph{consuming} a resource.
In general, \emph{consuming} a resource is equated with its 
\emph{syntactic occurrence} in the program. However, as this section
makes clear, when considering linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following program in a Haskell-like language there are two
syntactic occurrences of \emph{handle}, and a let-bound computation that closes
this handle is bound to \lstinline{x} before the \lstinline{handle} is
returned. However, if the program is evaluated lazily, the handle is \emph{not}
closed because the computation that closes the handle is never evaluated (since
\lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively consumed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
Under the lens of \emph{non-strict} evaluation, this means that a syntactic
occurrence of a resource in a let binding only entails consuming that resource
iff the binding itself is consumed.
%
We argue that a linear type system for a language with non-strict evaluation
semantics should accept the above program, unlike a linear type system for the
same program if it were to be evaluated \emph{eagerly}.
%
From this observation, and exploring the connection between computation and evaluation,
% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
% above example and for programs in general, should be defined in function of the
% language's evaluation strategy.
%
we turn our focus to \emph{linearity} under \emph{call-by-need}, not only
because GHC Core is \emph{call-by-need}, but also because the distinction
between semantically and syntactically consuming a resource is only exposed
under \emph{non-strict} semantics.
%
% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.

@@

Aligned with our original motivation of typechecking linearity in GHC Core such
that optimising transformations preserve linearity, and with the goal
of understanding linearity in a non-strict context, this section aims to
build an intuition for semantic linearity through examples of Core
programs that are semantically linear but are rejected by (the current) Core linear type
system.
% BETTER?:
%Aligned with our motivation of typechecking linearity in GHC Core such
%that optimising transformations preserve linearity, and in order to explain
%linearity in a non-strict context, we illustrate semantic linearity
%through examples of Core programs that are semantically linear but rejected by
%Core's linear type system -- building intuition towards semantic linearity in a
%lazy language.
%
In the examples, a \colorbox{working}{\workingcolorname} background highlights programs that are
syntactically linear and accepted by Core's naive linear type system. A
\colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
\emph{semantically} linear, but are not seen as linear by Core's (naive wrt lazy evaluation) linear type
system. Notably, the linear type system we develop in this work accepts all
\colorbox{notyet}{\notyetcolorname} programs.
% (while the few \colorbox{limitation}{\limitationcolorname} programs are not accepted).
A \colorbox{noway}{\nowaycolorname} background indicates that the program
violates linearity semantically, i.e.~the program effectively
discards or duplicates linear resources. We discuss the three main
syntactic constructs of Core, let bindings (non-recursive and
recursive) and case expressions.
